Pool Model - A unified Mathematical\\ Modelling Framework for Food Science

### 2025-11-04 paper/pool_model.tex
relabel supplementary sections
remove code samples

\setlength{\tabcolsep}{1.3pt}
\setlength{\columnsep}{0.4cm}
\usepackage{acronym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{bm}
\usepackage{booktabs}
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off, font=small]{caption}
\usepackage{diagbox}
\usepackage{float}
\usepackage[T1]{fontenc}
% \usepackage[a4paper,inner=2cm,outer=2cm,top=2.5cm,bottom=2.5cm]{geometry}
\usepackage{graphicx}% Include figure files
\usepackage{here}
\usepackage{hyperref}
\usepackage{lipsum}
\usepackage{listings}
\usepackage{listings-rust}
\usepackage[switch]{lineno}
\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{multirow}
\usepackage{newfloat}
\usepackage{onimage}
\usepackage{siunitx}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{tikz}
%\usepackage[dvipsnames]{xcolor}
\usepackage{xpatch}

@@

% Define Rust lisintgs
\lstset{
    language=Rust,
    showstringspaces=false
}
\DeclareFloatingEnvironment[
  name=Listing,
  listname=List of Listings,
  within=section,
]{listing}

@@

\newcommand{\beginsupplement}{%
  \renewcommand{\thesection}{}
  \setcounter{subsection}{0}
  \renewcommand{\thesubsection}{S\arabic{subsection}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
  \setcounter{listing}{0}
  \renewcommand{\thelisting}{S\arabic{listing}}
}
\newcommand{\mbx}{\mathbf{x}}
\newcommand{\mbm}{\mathbf{m}}
\newcommand{\mbp}{\mathbf{p}}
\newcommand{\mbu}{\mathbf{u}}
\newcommand{\mbF}{\mathbf{F}}
\newcommand{\mbH}{\mathbf{H}}
\newcommand{\cmt}[1]{\textcolor{red} {[#1]}}
\newcommand{\etal}{{\textit{et al. }}}

@@

\lstdefinelanguage{json}{
    basicstyle=\normalfont\ttfamily,
    numbers=left,
    numberstyle=\scriptsize,
    stepnumber=1,
    numbersep=8pt,
    showstringspaces=false,
    breaklines=true,
    frame=lines,
    backgroundcolor=\color{background},
    literate=
     *{0}{{{\color{numb}0}}}{1}
      {1}{{{\color{numb}1}}}{1}
      {2}{{{\color{numb}2}}}{1}
      {3}{{{\color{numb}3}}}{1}
      {4}{{{\color{numb}4}}}{1}
      {5}{{{\color{numb}5}}}{1}
      {6}{{{\color{numb}6}}}{1}
      {7}{{{\color{numb}7}}}{1}
      {8}{{{\color{numb}8}}}{1}
      {9}{{{\color{numb}9}}}{1}
      {:}{{{\color{punct}{:}}}}{1}
      {,}{{{\color{punct}{,}}}}{1}
      {\{}{{{\color{delim}{\{}}}}{1}
      {\}}{{{\color{delim}{\}}}}}{1}
      {[}{{{\color{delim}{[}}}}{1}
      {]}{{{\color{delim}{]}}}}{1},
}

@@

\subsubsection*{Deriving the pool model}
Since the \ac{abm} can be considered more fundamental than the \ac{ode} model, we should be able to derive the latter one from the former.
This can be achieved by starting with equations~\eqref{eq:spatial-full-system-simplified}.
We integrate the spatial equations over the total domain starting with the resource concentration $[R]$.
\begin{alignat}{7}
    \int\limits_D\dot{[R]}\text{dVol} 
    &= \int\limits_D\bigg[&&\sum\limits_c &&-\frac{u}{1+\nu_I [I]} \frac{V_c}{V_D} [R] \delta(x-x_c) &&+ D_R\Delta [R]\bigg]\text{dVol}\\
    &= \int\limits_D\bigg[&&\sum\limits_c &&-\frac{u}{1+\nu_I [I]} \frac{V_c}{V_D} [R] \delta(x-x_c)\bigg]\text{dVol} &&+ \int\limits_D D_R\Delta [R]\text{dVol}\\
    &= \int\limits_D\bigg[&&\sum\limits_c &&-\frac{u}{1+\nu_I [I]} \frac{V_c}{V_D} [R] \delta(x-x_c)\bigg]\text{dVol} &&+ \int\limits_{\partial D} D_R\vec{\nabla} [R]\vec{\text{dA}}\\
    &= &&\sum\limits_c &&-\frac{u}{1+\nu_I [I]} V_c \int\limits_D \frac{1}{V_D} [R] \delta(x-x_c)\text{dVol}\\
    &= &&\sum\limits_c &&-\frac{u}{1+\nu_I [I]} V_c [R]_{x=x_c}\\
    &= &&\sum\limits_c &&-\frac{u}{1+\nu_I [I]} V_c [R]\\
    &= &&&&-\frac{u}{1+\nu_I [I]V_D V_D^{-1}} V [R]\frac{V_D}{V_D}\\
    \dot{R} &= &&&&-\frac{u}{1+\nu_I V_D^{-1}I} \frac{1}{V_D} R V\\
We can split up the integral into two parts.
The second part can then be transformed into an integral over the flux of the resource along the boundary.
Since we chose Neumann boundary conditions, this contribution vanishes identically.
Afterwards, the first part can be evaluated by using the delta-distribution.
Note that we additionally need to consume a $V_D^{-1}$ term since our volume is finite.
We evaluate the resource concentration $[R]$ at the position of the individual cells and sum over all contributions.
Now we assume homogeneity in the sense, that we can reduce $[R]$ to only a single variable instead of considering its spatial distribution.
This means that every contribution does not depend on the position at which it was evaluated, i.e., $[R]_{x=x_c}=[R]$.
Finally, we can perform the sum to gain the total volume of all cells $V=\sum\limits_c V_c$.
\begin{alignat}{7}
    \int\limits_D\dot{[I]} &= \int\limits_D \bigg[&&\sum\limits_c b\frac{V_c}{V_D}&&\delta(x-x_c) &&+ D_I\Delta [I]\bigg]\text{dVol}\\
    &= \int\limits_D&&\sum\limits_c b\frac{V_c}{V_D}&&\delta(x-x_c) \text{dVol} &&+ \int\limits_D D_I\Delta [I]\text{dVol}\\
    &= \int\limits_D&&\sum\limits_c b\frac{V_c}{V_D}&&\delta(x-x_c) \text{dVol} &&+ \int\limits_{\partial D} D_I\nabla [I]\vec{\text{dA}}\\
    &= &&\sum\limits_c b V_c\int\limits_D \frac{1}{V_D} &&\delta(x-x_c) \text{dVol}\\
    \dot{I} &= b V
Following the same principle as before, we can integrate the inhibitor equation.
For the equation describing the growth of the cell volume, we simply add up all cells to obtain the total volume of all cells.
    \sum\limits_c\dot{V_c}
    &= \sum\limits_c&&\frac{u}{1+\nu_I [I]} \sigma [R]V_c\\
    &= &&\frac{u}{1+\nu_I [I]} \sigma [R]V\\
    \dot{V} &= &&\frac{u}{1+\nu_I V_D^{-1} I} \frac{1}{V_D}\sigma R V
We can thus summarize the equations obtained
It is clear that the original model is recovered this way.
To obtain the full equations, we additionally need to consider transition between states of lag-phase and growth phase not depicted here.
The preceding equations all assume that cells are in growth phase.
It is however straightforward, to set up these equations.
\subsubsection{Code samples}
The code samples show sample code from the simulation that calculate cellular interactions and cell cycle.
Since backend of the simulation is written in Rust, so is the implementation of cellular aspects.
Additionally, bindings are generated such that the full power of the simulation tool can be accessed.
This significantly improves handling and development time of new models.
\begin{listing}
\begin{lstlisting}[language=rust]
let z = ext_pos - own_pos;
let r = z.norm();
let sigma = r / (self.cell_radius + ext_radius);
if sigma < 1.0 {
    let q = 0.2;
    let dir = z.normalize();
    let modifier = (1.0 + q) / (q + sigma);
    return self.potential_strength * dir * modifier;
} else {
    return Vector2::zero();
\end{lstlisting}
    Sample code for calculation of interactions between two bacteria cells.
\label{listing:spatial-interaction-strength}
\end{listing}
\begin{listing}
\begin{lstlisting}[language=rust]
// Clone existing cell
let mut c2 = c1.clone();

let r = c1.interaction.cell_radius;

// Make both cells smaller
c1.interaction.cell_radius /= std::f64::consts::SQRT_2;
c2.interaction.cell_radius /= std::f64::consts::SQRT_2;

// Generate cellular splitting direction randomly
let angle_1 = rng.gen_range(0.0..2.0 * std::f64::consts::PI);
let dir_vec = nalgebra::Rotation2::new(angle_1) * Vector2::from([1.0, 0.0]);

// Define new positions for cells
// It is randomly chosen if the old cell is left or right
let offset = dir_vec * r / std::f64::consts::SQRT_2;
let old_pos = c1.pos();

// Increase the remaining food in the bacteria by the fraction of volumes
// from before and after the division. This exactly coincides with factor 2
// since we half the volume of the cell
c1.cellular_reactions.intracellular_concentrations[0] *= 2.0;
c2.cellular_reactions.intracellular_concentrations[0] *= 2.0;

c1.set_pos(&(old_pos + offset));
c2.set_pos(&(old_pos - offset));
\end{lstlisting}
\caption{
    Sample code for dividing one bacteria cell into two.
}
\end{listing}
%
%
% ................................................................................................
\subsubsection{Full list of parameters}
\label{sec:supplement-full-parameters}

>>>>

\setlength{\tabcolsep}{1.3pt}
\setlength{\columnsep}{0.4cm}
\usepackage{acronym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{bm}
\usepackage{booktabs}
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off, font=small]{caption}
\usepackage{diagbox}
\usepackage{float}
\usepackage[T1]{fontenc}
% \usepackage[a4paper,inner=2cm,outer=2cm,top=2.5cm,bottom=2.5cm]{geometry}
\usepackage{graphicx}% Include figure files
\usepackage{here}
\usepackage{hyperref}
\usepackage{lipsum}
\usepackage[switch]{lineno}
\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{multirow}
\usepackage{newfloat}
\usepackage{onimage}
\usepackage{siunitx}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{tikz}
%\usepackage[dvipsnames]{xcolor}
\usepackage{xpatch}

@@

@@

\newcommand{\beginsupplement}{%
  \renewcommand{\thesection}{}
  \setcounter{section}{0}
  \renewcommand{\thesection}{S\arabic{section}}
  % \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  % \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
}
\newcommand{\mbx}{\mathbf{x}}
\newcommand{\mbm}{\mathbf{m}}
\newcommand{\mbp}{\mathbf{p}}
\newcommand{\mbu}{\mathbf{u}}
\newcommand{\mbF}{\mathbf{F}}
\newcommand{\mbH}{\mathbf{H}}
\newcommand{\cmt}[1]{\textcolor{red} {[#1]}}
\newcommand{\etal}{{\textit{et al. }}}

@@

@@

\subsubsection*{Deriving the pool model}
Since the \ac{abm} can be considered more fundamental than the \ac{ode} model, we should be able to derive the latter one from the former.
This can be achieved by starting with equations~\eqref{eq:spatial-full-system-simplified}.
We integrate the spatial equations over the total domain starting with the resource concentration $[R]$.
\begin{alignat}{7}
    \int\limits_D\dot{[R]}\text{dVol} 
    &= \int\limits_D\bigg[&&\sum\limits_c &&-\frac{u}{1+\nu_I [I]} \frac{V_c}{V_D} [R] \delta(x-x_c) &&+ D_R\Delta [R]\bigg]\text{dVol}\\
    &= \int\limits_D\bigg[&&\sum\limits_c &&-\frac{u}{1+\nu_I [I]} \frac{V_c}{V_D} [R] \delta(x-x_c)\bigg]\text{dVol} &&+ \int\limits_D D_R\Delta [R]\text{dVol}\\
    &= \int\limits_D\bigg[&&\sum\limits_c &&-\frac{u}{1+\nu_I [I]} \frac{V_c}{V_D} [R] \delta(x-x_c)\bigg]\text{dVol} &&+ \int\limits_{\partial D} D_R\vec{\nabla} [R]\vec{\text{dA}}\\
    &= &&\sum\limits_c &&-\frac{u}{1+\nu_I [I]} V_c \int\limits_D \frac{1}{V_D} [R] \delta(x-x_c)\text{dVol}\\
    &= &&\sum\limits_c &&-\frac{u}{1+\nu_I [I]} V_c [R]_{x=x_c}\\
    &= &&\sum\limits_c &&-\frac{u}{1+\nu_I [I]} V_c [R]\\
    &= &&&&-\frac{u}{1+\nu_I [I]V_D V_D^{-1}} V [R]\frac{V_D}{V_D}\\
    \dot{R} &= &&&&-\frac{u}{1+\nu_I V_D^{-1}I} \frac{1}{V_D} R V\\
\end{alignat}
We can split up the integral into two parts.
The second part can then be transformed into an integral over the flux of the resource along the boundary.
Since we chose Neumann boundary conditions, this contribution vanishes identically.
Afterwards, the first part can be evaluated by using the delta-distribution.
Note that we additionally need to consume a $V_D^{-1}$ term since our volume is finite.
We evaluate the resource concentration $[R]$ at the position of the individual cells and sum over all contributions.
Now we assume homogeneity in the sense, that we can reduce $[R]$ to only a single variable instead of considering its spatial distribution.
This means that every contribution does not depend on the position at which it was evaluated, i.e., $[R]_{x=x_c}=[R]$.
Finally, we can perform the sum to gain the total volume of all cells $V=\sum\limits_c V_c$.
\begin{alignat}{7}
    \int\limits_D\dot{[I]} &= \int\limits_D \bigg[&&\sum\limits_c b\frac{V_c}{V_D}&&\delta(x-x_c) &&+ D_I\Delta [I]\bigg]\text{dVol}\\
    &= \int\limits_D&&\sum\limits_c b\frac{V_c}{V_D}&&\delta(x-x_c) \text{dVol} &&+ \int\limits_D D_I\Delta [I]\text{dVol}\\
    &= \int\limits_D&&\sum\limits_c b\frac{V_c}{V_D}&&\delta(x-x_c) \text{dVol} &&+ \int\limits_{\partial D} D_I\nabla [I]\vec{\text{dA}}\\
    &= &&\sum\limits_c b V_c\int\limits_D \frac{1}{V_D} &&\delta(x-x_c) \text{dVol}\\
    \dot{I} &= b V
\end{alignat}
Following the same principle as before, we can integrate the inhibitor equation.
For the equation describing the growth of the cell volume, we simply add up all cells to obtain the total volume of all cells.
\begin{alignat}{3}
    \sum\limits_c\dot{V_c}
    &= \sum\limits_c&&\frac{u}{1+\nu_I [I]} \sigma [R]V_c\\
    &= &&\frac{u}{1+\nu_I [I]} \sigma [R]V\\
    \dot{V} &= &&\frac{u}{1+\nu_I V_D^{-1} I} \frac{1}{V_D}\sigma R V
\end{alignat}
We can thus summarize the equations obtained
\begin{align}
    \dot{R} &= -\frac{u}{1+\nu_I V_D^{-1}I} \frac{1}{V_D} R V\\
    \dot{I} &= b V\\
    \dot{V} &= \frac{u}{1+\nu_I V_D^{-1} I} \frac{1}{V_D}\sigma R V.
\end{align}
It is clear that the original model is recovered this way.
To obtain the full equations, we additionally need to consider transition between states of lag-phase and growth phase not depicted here.
The preceding equations all assume that cells are in growth phase.
It is however straightforward, to set up these equations.
%
%
% ................................................................................................
\subsubsection{Full list of parameters}
\label{sec:supplement-full-parameters}
